---
title: "Reproducible Research: Peer Assessment 1"
author: "Frank Hecker"
output:
  html_document:
    keep_md: true
    pandoc_args: [
      "+RTS", "-K32m", "-RTS"
    ]
---

## Introduction

I carried out the analysis for this project using
`r version$version.string`; I used the
[dplyr](http://cran.r-project.org/web/packages/dplyr) package for data
analysis and the
[ggplot2](http://cran.r-project.org/web/packages/ggplot2) package for
plotting. I also used the `md5sum` function in the tools package in
core R, and the `date_format` function in the
[scales](http://cran.r-project.org/web/packages/scales) package. For
complete information on the R environment I used please see Appendix
2 below.

I first loaded the additional libraries needed for this project; the
dplyr package masks some functions from other packages, so I
suppressed any warnings resulting from such conflicts. Also, the plyr
and dplyr packages can conflict with one another, so I first unloaded
the plyr package if it was loaded.

```{r load_data}
if ("package:plyr" %in% search()) detach("package:plyr", unload = TRUE)
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
library("tools")
library("scales")
```

I used the dplyr package because it provides a consistent vocabulary
for actions taken during analysis and for the overall processing
pipeline. For a quick overview see "[Introduction to
dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)".
Note that the idiom `d %>% f(x)` is equivalent to `f(d, x)`; see the
section "Chaining" in "Introduction to dplyr".

## Loading and preprocessing the data

The raw data for this project was included in Roger Peng's original
repository in the form of a file `activity.zip`. In order to ensure
that I was working with the correct data, I first computed the MD5
hash of this file and compared it to the known value I previously
computed against the file as cloned from the original GitHub
repository.

```{r check_md5sum}
stopifnot(md5sum("activity.zip") == "61e0c1f8c4a736ff1ab0f8863344b301")
```

I then extracted the CSV file `activity.csv` and read the data into a
data frame:

```{r extract_csv}
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
str(activity)
```

As displayed by the `str` function, the `activity` data frame contains
`r nrow(activity)` observations of three variables:

* `date`: The date of each observation, in the form "YYYY-MM-DD".
* `interval`: The 5-minute interval during the day during which the
  number of steps was recorded, expressed as 100 * H + M, where H
  is the hour from 0 to 23, and M is the starting minute of the
  interval within the hour, from 0 to 55.
* `steps`: The number of steps recorded during the given interval
  (missing for some observations and recorded as NA).

In order to carry out the analysis of weekdays versus weekends below,
I also converted the `date` variable to POSIX time format:

```{r date_convert}
activity <- activity %>%
        mutate(date = as.POSIXct(date, tz = "GMT"))
str(activity)
```

## What is the mean total number of steps taken per day?

In order to present a histogram of the data I first calculated the
total number of steps for each day, ignoring missing values:

```{r steps_per_day}
dailySteps <- activity %>%
        group_by(date) %>%
        summarise(steps = sum(steps, na.rm = TRUE))
str(dailySteps)
```

I then plotted a histogram to show how many days the subject walked a
certain number of steps; I set the limits on the y axis to be
consistent with those on a second histogram below:

```{r steps_histogram}
p <- ggplot(dailySteps, aes(x = steps))
p <- p + geom_histogram(stat = "bin", binwidth = 1000)
p <- p + coord_cartesian(ylim=c(0, 20))
p <- p + xlab("Total Steps Per Day")
p <- p + ylab("Number of Days")
print(p)
```

Several days recorded zero or very few steps; this could be either
because the subject didn't walk much at all that day, or because few
or no values were recorded for that day. (Since I ignored missing
values in computing the sum above, a day for which all intervals had
values of NA would result in a sum of zero steps.)

I then calculated the mean and median; using `na.rm = TRUE` was not
really necessary since based on the above computation
`dailySteps$steps` would always have a value (which might be zero if
all `steps` values for a day were missing):

```{r mean_median}
mean(dailySteps$steps, na.rm = TRUE)
median(dailySteps$steps, na.rm = TRUE)
```

Although the histogram's peak was around 10,000 steps per day the mean
was brought down because of the large number of days with zero (or
almost zero) steps. The median is less sensitive to this effect.

## What is the average daily activity pattern?

Each day's observations reflect the subject's activity during that
day. To get a sense for what an average day looks like, I grouped the
observations for each 5-minute interval and calculated the average
number of steps for that interval across all days (ignoring any
missing values).

```{r avgActivity}
avgActivity <- activity %>%
        group_by(interval) %>%
        summarise(steps = mean(steps, na.rm = TRUE))
str(avgActivity)
```

I then plotted the average number of steps for each interval:

```{r avgActivity_plot}
p <- ggplot(avgActivity, aes(x = interval, y = steps))
p <- p + geom_line()
p <- p + xlab("5-Minute Interval")
p <- p + ylab("Steps")
p <- p + ggtitle("Average Daily Activity")
print(p)
```

From the above graph the interval with the most number of steps on
average appeared to be somewhere between 800 and 900. I used the
`top_n` function of the dplyr package to determine the specific
interval with the highest average number of steps:

```{r top_interval}
top_n(avgActivity, 1, steps)
```

See Appendix 1 for an alternative way to plot this data.

## Imputing missing values

Out of the `r nrow(activity)` rows in the `activity` data frame, I
calculated the number of missing values (rows with an NA value for
`steps`) as follows:

```{r missing_values}
sum(is.na(activity$steps))
```

(Although it's not required for the project, I also found the number
of days for which all values were missing, by counting the number of
observations for each day that had a valid value for `steps` and then
counting the number of days with zero valid observations.)

```{r missing_days}
missingDays <- activity %>%
        group_by(date) %>%
        summarise(validObs = sum(!is.na(steps))) %>%
        filter(validObs == 0) %>%
        nrow()
missingDays
```

In cases where a value was missing for a particular interval I chose
to impute the missing value using the average number of steps for that
interval across all days.

In particular I used the `mutate` function of the dplyr package to
change the values of the `steps` variable for all observations in the
`activity` data frame. If the existing value of the `steps` variable
was NA then I set the new value to be the mean of all values of
`steps` for that interval (since I grouped based on the `interval`
variable); otherwise I set the new value to be the same as the
original value.

```{r impute_missing}
imputed <- activity %>%
        group_by(interval) %>%
        mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
```

I then recalculated the average number of steps for each interval
using the `imputed` data instead of the original `activity` data;
otherwise the calculation was exactly the same:

```{r imputed_avg}
dailyImputed <- imputed %>%
        group_by(date) %>%
        summarise(steps = sum(steps, na.rm = TRUE))
```

I then replotted the corresponding histogram:

```{r imputed_histogram}
p <- ggplot(dailyImputed, aes(x = steps))
p <- p + geom_histogram(stat = "bin", binwidth = 1000)
p <- p + coord_cartesian(ylim=c(0, 20))
p <- p + xlab("Total Steps Per Day")
p <- p + ylab("Number of Days")
p <- p + ggtitle("Average Daily Steps (Missing Values Imputed)")
print(p)
```

This new histogram appeared to be identical to the previous
histogram, except that the number of days with 0-999 steps decreased
from 10 to 2, while the number of days with 10,000-10,999 steps
increased from 10 to 18.

Finally, I calculated the mean and median of the new dataset with
imputed data:

```{r imputed_mean}
mean(dailyImputed$steps, na.rm = TRUE)
median(dailyImputed$steps, na.rm = TRUE)
```

This mean was higher, since I had imputed values for the
`r missingDays` days with no valid observations, days that had
previously contributed zero values to the mean.

Also, this median was the same as the mean. That's because the values
for every interval on the missing days had been replaced with values
corresponding to the average of all such intervals, and thus those
days had a correspondingly average total number of steps.

When sorting the days according to the total number of steps, the days
with all imputed values ended up in the middle of the list, and thus
the values of those days for the total number of steps per day was
chosen as the median number of steps per day. I confirmed this by
looking at the set of observations around the midpoint:

```{r midpoint_check}
midpoint <- nrow(dailyImputed) %/% 2
midpoint
dailyImputed <- dailyImputed %>% arrange(steps)
dailyImputed[(midpoint - 6):(midpoint + 6), ]
```

The entries with value equal to the mean correspond to the days for
which all values were missing and were imputed.

## Are there differences in activity patterns between weekdays and weekends?

To analyze the differences between activity patterns on weekdays
versus weekends I first tagged all observations according to whether
they were recorded on weekdays or on weekends. (This was possible
since I had previously converted the `date` variable in the `activity`
data frame to have POSIX date values.)

```{r tag_days}
weekendDays <- c("Saturday", "Sunday")
activity <- activity %>%
        mutate(period = as.factor(ifelse(weekdays(date) %in% weekendDays,
                                         "weekend", "weekday")))
str(activity)
```

I then computed the average activity for weekends and for
weekdays. This was similar to how I previously computed the average
activity for all days, except that I grouped the data by both `period`
and `interval` in order to create separate sets of interval means for
weekdays versus weekends.

```{r avgActivityByPeriod}
avgActivityByPeriod <- activity %>%
        group_by(period, interval) %>%
        summarise(steps = mean(steps, na.rm = TRUE))
glimpse(avgActivityByPeriod)
```

I then plotted the data in a panel plot containing graphs for both the
weekday and weekend data:

```{r avgActivityByPeriod_plot}
p <- ggplot(avgActivityByPeriod,
            aes(x = interval, y = steps))
p <- p + facet_grid(period ~ .)
p <- p + geom_line()
p <- p + xlab("Interval")
p <- p + ylab("Steps")
p <- p + ggtitle("Average Daily Activity on Weekdays and Weekends")
print(p)
```

See Appendix 1 for an alternative way to plot this data.

## Appendix 1

Although it was not required for the project, I decided to create
alternative plots for the daily activity and daily activity by
weekends versus weekdays, in order to make the graphs more
understandable, as well as more correct. In particular, the interval
values in the above graphs were not evenly spread out on the x axis;
for example, for the first hour of the day the interval values went
from 45 to 50 to 55, and then jumped straight to 100. This led to
corresponding jumps in the plotted lines.

To create these alternative plots I used the interval values to
compute actual times in POSIX format. This required first converting
the interval values to seconds since midnight, and then converting the
seconds values to POSIX times on an arbitrarily-chosen date. (This
date was independent of the dates of the observations, and was ignored
for purposes of plotting; it was there only because a POSIX time has
to include a date component.)

I first added POSIX time information for the data frame containing the
average daily activity for all days combined:

```{r avgActivity_time}
avgActivity <- avgActivity %>%
        mutate(time = 3600 * interval %/% 100 + 60 * interval %% 100) %>%
        mutate(time = as.POSIXct(time, origin = "1960-01-01", tz = "GMT"))
str(avgActivity)
```

The x axis of the resulting plot showed the actual times of day for
each interval in terms of hours and minutes:

```{r avgActivity_time_plot}
p <- ggplot(avgActivity, aes(x = time, y = steps))
p <- p + geom_line()
p <- p + scale_x_datetime(labels = date_format(format = "%H:%M"))
p <- p + xlab("Time of Day")
p <- p + ylab("Steps Per 5-Minute Interval")
p <- p + ggtitle("Average Daily Activity")
print(p)
```

On a typical day the subject's activity rapidly increased when they
wake up around 6 am, reached a peak around 8:30 am, and then
fluctuated through the remainder of the morning and afternoon before
declining past 9 pm as bedtime approached.

The second plot was for average daily activity on weekdays versus
weekends. Again I converted the intervals to times of day, this time
in the data frame containing the tagged average activity data:

```{r avgActivityByPeriod_time}
avgActivityByPeriod <- avgActivityByPeriod %>%
        mutate(time = 3600 * interval %/% 100 + 60 * interval %% 100) %>%
        mutate(time = as.POSIXct(time, origin = "1960-01-01", tz = "GMT"))
glimpse(avgActivityByPeriod)
```

I then plotted the data with an appropriate x axis:

```{r avgActivityByPeriod_time_plot}
p <- ggplot(avgActivityByPeriod,
            aes(x = time, y = steps))
p <- p + facet_grid(period ~ .)
p <- p + geom_line()
p <- p + scale_x_datetime(labels = date_format(format = "%H:%M"))
p <- p + xlab("Time of Day")
p <- p + ylab("Steps Per 5-Minute Interval")
p <- p + ggtitle("Average Daily Activity on Weekdays and Weekends")
print(p)
```

On weekends the subject's activity didn't increase significantly until
after 8 am (versus about 6 am on weekdays), presumably because they
woke up later than on weekdays. During the day the subject was more
active on weekends than on weekdays. Finally, on weekends the
subject's activity peaked later in the evening, presumably reflecting
a later bedtime.

## Appendix 2

I used the following R environment in doing the analysis for this
project:

```{r sessionInfo}
sessionInfo()
```
